---
title: "3.3_apply"
format:
   revealjs:
     margin: 0.1
editor: visual
---

## `*apply` family

Imagine you have a list of vectors and you want to apply the same function(s) to each element of the list:

-   I apply the function manually, selecting the elements

-   a **`for`** loop that iterates over the elements of the list and applies the function(s)

    ...

```{r, echo=TRUE}
my_list=list(
    var1=rnorm(n = 100, mean = 0, sd = 1),
    var2=runif(n = 100, min = 0, max = 1),
    var3=rnorm(n = 100, mean = 0, sd = 1),
    var4=rnorm(n = 100, mean = 0, sd = 1)
)
```

## `*apply` family

::::: columns
::: {.column width="60%"}
Let's compute mean, median and std

```{r, echo=TRUE}
# initialize the vectors with arbitrary numbers,
# but as many as we need
means = 1:length(my_list)
medians = 1:length(my_list)
stds = 1:length(my_list)

# Loop
for(i in 1:length(my_list)){

  means[i] = mean(my_list[[i]])
  medians[i] = median(my_list[[i]])
  stds[i] = sd(my_list[[i]])
}
```
:::

::: {.column width="40%"}
Result

```{r, echo=TRUE}
means

medians

stds
```
:::
:::::

## `*apply` family

Everything works! but:

-   writing the for loop is laborious, with indices both for the list and for the vector we are filling

-   we need to pre-allocate variables (for speed reasons)

-   8 lines of code (for this simple example)

## `*apply` family

R has a family of apply functions such as **`lapply`**, **`sapply`**, etc. that let us get the same result in a more concise, fast and simple way:

```{r, echo=TRUE}
means = sapply(my_list, mean)
medians = sapply(my_list, median)
stds = sapply(my_list, sd)

means
medians
stds
```

## `*apply` family

**apply( \< list \> , \< function \> )**

-   What can the **list** be?
    -   a list

    -   a dataframe

    -   a vector
-   What can the **function** be?
    -   base or imported from a package

    -   custom

    -   anonymous

## `*apply` family

Before analysing the **`*apply`** family, I think a further parallel with the for loop we have seen is useful. **`apply`** is nothing more than a slightly simplified **`for`** loop.

::::: columns
::: {.column width="50%"}
**`for`** loop

```{r, echo=TRUE}
vec = 1:5
for(i in vec){
  print(i)}
```
:::

::: {.column width="50%"}
**`sapply`**

```{r, echo=TRUE}
vec = 1:5
res = sapply(vec, print)
```
:::
:::::

## `*apply` family - custom function

We can also use functions we created ourselves:

```{r, echo=TRUE}
center_var = function(x){
  x - mean(x)
}

my_list = list(
    vec1 = runif(n = 10, min  = 0, max = 1),
    vec2 = runif(n = 10, min  = 0, max = 1),
    vec3 = runif(n = 10, min  = 0, max = 1)
)

lapply(my_list, center_var)
```

## `*apply` family - implicit vs. explicit

So, just as the **`for`** loop written as **i** in **vec** assigns to **i** one **element** of the **vec** object at a time, internally the **`*apply`** functions take the **first element** of the **input** object (list) and directly **apply** the function we have chosen.

::::: columns
::: {.column width="40%"}
implicit **`sapply`**

```{r, echo=TRUE}
vec = 1:5
res = sapply(vec, print)
```
:::

::: {.column width="60%"}
explicit **`sapply`**

```{r, echo=TRUE}
vec = 1:5
res = sapply(vec, function(i) print(i))
```
:::
:::::

## `*apply` family - anonymous function

An anonymous function is a function that is not saved in an object but written to be **executed directly**, inside other functions that allow it:

```{r, echo=TRUE}
lapply(my_list, function(x) x - mean(x))
```

**`x`** is just a **placeholder** (analogous to i) and can be any letter or name!

## All the types of `*apply`

-   **`lapply()`**: the basic function

-   **`sapply()`**: simplified-apply

-   **`tapply()`**: rarely used, useful with factors

-   **`apply()`**: useful for dataframes/matrices

-   **`mapply()`**: multivariate version, uses several lists at the same time

-   **`vapply()`**: used inside functions and packages

## `lapply`

**`lapply`** stands for list-apply and always returns a list, applying the function to each element of the input list:

```{r, echo=TRUE}
res = lapply(my_list, mean)
res
class(res)
```

## `sapply`

**`sapply`** stands for simplified-apply and (tries to) return a simpler version of a list, applying the function to each element of the input list:

```{r, echo=TRUE}
res = sapply(my_list, mean)
res
class(res)
```

## `apply`

**`apply`** works specifically on dataframes or matrices, applying a function to the rows or the columns:

```{r, echo=TRUE}
my_df = data.frame(x1 = runif(n = 5,min = 1,max = 10),
                  x2 = runif(n = 5,min = 3,max = 4))
my_df
```

## `apply`

Apply the mean function to all rows (1):

```{r, echo=TRUE}
apply(my_df, MARGIN = 1, FUN = mean)
```

Apply the mean function to all columns (2):

```{r, echo=TRUE}
apply(my_df, MARGIN = 2, FUN = mean)
```

## `tapply`

**`tapply`** lets us apply a function to a vector, splitting the vector according to a categorical variable:

```{r, echo=TRUE}
vec = rnorm(75)

index = rep(c("a", "b", "c"), each = 25)

tapply(vec, INDEX = index, mean)
```

Here *`INDEX`* is a character vector or a factor.

## `tapply`

In this case we compute the mean for each age category:

```{r, echo=TRUE, message=FALSE, warning=FALSE}
my_df = readr::read_csv("data/mydf_2.csv")
head(my_df)
tapply(my_df$age, my_df$age_cat, mean)
```

## `vapply`

**`vapply`** is a more robust version of the previous ones from a programming point of view. In practice it lets (and requires) us to specify in advance the type of data we expect as a result:

```{r, echo=TRUE}
vapply(my_list, FUN = mean, FUN.VALUE = numeric(length = 1))
```

*`FUN.VALUE = numeric(length = 1)`*: indicates that each result is a single numeric value.

## `mapply`

**`mapply`** lets us handle **several lists at the same time** for more complex scenarios. For example, we want to use the **`rnorm()`** function to generate 4 samples combining different means and standard deviations:

```{r, echo=TRUE}
mus = list(10, 20, 30, 40)
stds = list(1, 2, 3, 4)

mapply(function(x,y) rnorm(n = 5, mean = x, sd = y),
       mus, stds, SIMPLIFY = FALSE)
```

**IMPORTANT**: all the lists included must have the same length!

## `mapply` {.smaller}

**mapply(function(x,y) rnorm(n = 5, mean = x, sd = y), mus, stds, SIMPLIFY = FALSE)**

-   **`function(...)`**: an anonymous function, as we saw before, which can have n elements

-   **`rnorm(n = 5, mean = x, sd = y)`**: the actual anonymous function, where we have the placeholders x and y

-   **`mus, stds`**: the lists corresponding, **in order**, to the placeholders, so x = mus and y = stds

-   **`SIMPLIFY = FALSE`**: simply says to return a list and not to try (like sapply) to simplify the result

## `mapply` as a `for`

The same result (in a more verbose way) can be obtained with a **`for`** loop, using the iterator i several times:

```{r, echo=TRUE}
mus = list(10, 20, 30)
stds = list(1,2,3)
res = vector(mode = "list", length = length(mus)) # empty list

for(i in 1:length(mus)){
  res[[i]] = rnorm(n = 6, mean = mus[[i]], sd = stds[[i]])
}
res
```

## `replicate`

This function lets us repeat an operation n times, without using an iterator or a placeholder.

**`replicate(n, expr)`**

-   ***`n`*** is the number of repetitions

-   ***`expr`*** is the portion of code to repeat

## `replicate`

We sample 1000 times from a normal distribution and compute the mean, AKA the sampling distribution of the mean

::::: columns
::: {.column width="50%"}
```{r, echo=TRUE}
nrep=1000

nsample=30

mu=100

sd=30

means=
  replicate(
    n = nrep,
    expr = {mean(
      rnorm(nsample, mu, sd))}
    )
```
:::

::: {.column width="50%"}
```{r, echo=FALSE}
hist(means, main = "Sampling distribution of the mean")
abline(v = mean(means), col = "blue", lwd = 3, lty = 2)  # Estimated mean
abline(v = c(mean(means) - sd(means), mean(means) + sd(means)), col = "red", lwd = 3, lty = 3)  # Standard error
```
:::
:::::

## Now let's practice! {style="text-align: center;"}

<br/> Open this link and keep it open:

<https://etherpad.wikimedia.org/p/arca-corsoR>
