---
title: "Factors"
format:
   revealjs:
     margin: .05
     fontsize: 1.9em
     linestretch: 1.4
     mainfont: Helvetica Neue
editor: visual
---

## Data Structures

-   vectors

-   ***`factors`***

-   lists

-   matrices

-   arrays

-   dataframes

## Factors

Factors are a peculiar data type: although they look like simple **`characters`**, they are actually a type of **`integer`** vector with some additional properties.

## Creating a factor

Factors can be created either by converting a character vector with the **`as.factor()`** command ...

```{r, echo=TRUE}
char_vect = rep(c("hello","ciao", "hola"), each = 2)
char_vect

my_fact = as.factor(char_vect)
my_fact
```

... or by explicitly creating a factor with the **`factor()`** command:

```{r, echo=TRUE}

my_fact = factor(x = char_vect)
my_fact
```

## Properties of factors

In practice, they assign a label to an integer value:

```{r, echo = TRUE}
typeof(my_fact) # typeof determines the type or storage mode of any object
```

```{r, echo=TRUE, warning=TRUE}
as.integer(my_fact)

my_fact
```

::: fragment
This does not hold for character vectors!

```{r, echo = TRUE}
typeof(char_vect) # typeof determines the type or storage mode of any object
```

```{r, echo = TRUE}
as.integer(char_vect)

char_vect
```
:::

------------------------------------------------------------------------

Factors store **`levels()`** as metadata,

```{r, echo = TRUE}
levels(my_fact)
```

::: {.fragment}
regardless of which ones are actually present in the vector. For example, if I create a factor made only of the elements of my_fact different from ciao:

```{r, echo = TRUE}
my_fact2 = my_fact[my_fact!="ciao"]
```

:::

::: {.fragment}
The levels of my_fact2 will be the same as those of my_fact (ciao included), even though ciao is no longer present as an observation:

```{r, echo = TRUE}
my_fact2
levels(my_fact2)
```
:::
------------------------------------------------------------------------

However, we can drop the levels that are no longer needed with the `droplevels()` command:

```{r, echo = TRUE}
# what would my_fact2 look like if I dropped the levels with no observations?
droplevels(my_fact2)
```

::: fragment
```{r, echo = TRUE}
# modify my_fact2 by removing the unused levels
my_fact2 = droplevels(my_fact2)
my_fact2
```
:::

------------------------------------------------------------------------

We can also **rename** the levels of a factor...

```{r, echo=TRUE}
my_fact # the factor I created
my_fact_lev = my_fact # create another identical factor "my_fact_lev"
```

::: fragment
by accessing them with the `levels` command:

```{r, echo=TRUE}
levels(my_fact_lev)
# rename the levels of the factor
levels(my_fact_lev) = c("italian","english","spanish")
```
:::

::: fragment
Now the two factors are:

```{r, echo=TRUE}
my_fact_lev
my_fact
```
:::

------------------------------------------------------------------------

We can **both reorder** and **rename** the levels of a factor with the `factor()` function:

```{r, echo=TRUE}
my_fact
# by default levels are assigned in alphabetical order
as.integer(my_fact)
```

::: fragment
We can decide that the first level should be hello:

```{r, echo=TRUE}
new_fact = factor(x = my_fact, levels = c("hello","ciao","hola"))
new_fact
as.integer(new_fact)
```
:::

------------------------------------------------------------------------

We can also rename the levels, again with the `factor()` function:

```{r, echo=TRUE}
my_fact

new_fact2 = factor(x = my_fact, levels = c("hello","ciao","hola"),
                   labels = c("english","italian","spanish"))
new_fact2
```

::: {.fragment}

Previous versions:

```{r, echo=TRUE}
new_fact

my_fact
```
:::

------------------------------------------------------------------------

**Remember**: the ***`levels`*** argument inside the **`factor()`** function is used to **reorder** the levels of the factor, while the **`levels()`** function is used to access the level names and, if needed, **rename** them!

```{r, echo=TRUE}
# the factor I created
my_fact
levels(my_fact)
levels(my_fact) = c("hello","ciao","hola")
```

::: fragment
What do you notice compared to the my_fact vector shown above?

```{r, echo=TRUE}
my_fact
as.integer(my_fact)
```
:::

## Ordinal Factors

A subtype of factors are ***ordered factors***, which correspond to ordinal variables, i.e. variables whose levels are sorted in increasing order:

```{r, echo=TRUE}
my_ord_fac = as.ordered(my_fact)

my_ord_fac # alphabetical order

attributes(my_ord_fac)
```

## Let's practice! {style="text-align: center;"}

<br/> Open this link and keep it open:

<https://etherpad.wikimedia.org/p/arca-corsoR>
