---
title: "Lists, Matrices and Arrays"
format:
   revealjs:
     margin: .05
     fontsize: 1.9em
     linestretch: 1.4
     mainfont: Helvetica Neue
editor: visual
---

## Data Structures

-   vectors

-   factors

-   ***`lists`***

-   matrices

-   arrays

-   dataframes

## Lists

Lists are flexible structures that can contain objects of different types and different sizes (each element can itself be a list), and they are created with the `list()` command

```{r, echo=TRUE}
my_list = list("hello!", 1:3, c(TRUE,FALSE),
               list("hello!", 1:3, c(TRUE,FALSE)))
str(my_list)
```

## Lists

Although a list is one-dimensional, it develops in depth

```{r, echo=TRUE}
str(my_list)
```

## Lists

***First Level***: list

```{r, echo=TRUE}
str(my_list[2]) # single brackets []
```

***Second Level***: numeric vector

```{r, echo=TRUE}
str(my_list[[2]]) # double brackets [[]]
```

***Third Level***: integer

```{r, echo=TRUE}
str(my_list[[2]][1]) # double brackets [[]] and chaining
```

## Lists

We can assign names to the elements of a list...

```{r, echo=TRUE}
my_list_nam = list(word = "hello!",
                   numbers =  1:3,
                   logical = c(TRUE,FALSE),
                   list = list("hello!", 1:3, c(TRUE,FALSE)))
my_list_nam
```

## Lists

and access them by calling them by "name" with the **`$`** operator

```{r, echo=TRUE}
my_list_nam$word

my_list_nam$list # list created as an element of the list
```

## Lists - Attributes

Like vectors, lists also have a length (**`length()`**) and possibly names (**`names()`**).

```{r, echo=TRUE}

length(my_list_nam)

names(my_list_nam)
```

::: fragment
The **`str()`** (structure) command is very useful for lists because it gives an overview of their structure:

```{r, echo=TRUE}
str(my_list_nam)
```
:::

## Lists - Indexing

The difference between the square brackets is whether we want to subset the list, getting another list, or directly access the inner element:

```{r, echo=TRUE}
my_list_nam[1] # getting another list

my_list_nam[[1]] # accessing the inner element
```

## Lists - Indexing

If we want to actually subset the list, we must always use single square brackets:

```{r, echo=TRUE}
# list made of the first two elements of my_list_nam
my_list2 = my_list_nam[1:2]
str(my_list2)
```

## Lists - Nested Indexing

Lists have a one-dimensional structure that can develop in depth. To select nested elements we can chain several brackets:

```{r, echo=TRUE}
my_list_nam[2]

my_list_nam[2][1] # wrong!

my_list_nam[[2]][1]

```

## Lists - Why?

They are flexible and allow us to store/organize complex and varied data/information. For example, the **`lm()`** function returns a list containing various pieces of information and results about the tested model.

```{r, echo=TRUE}
# Simulate 30 observations, assuming that the effect of x = 0.3
N = 30; b1 = 0.3; sigma = 1
x = rnorm(N, 0, 1); # independent variable
y = b1*x + rnorm(n = N, mean = 0, sd = sigma) # effect plus error
```

::: fragment
```{r, echo=TRUE}
# linear model
mod = lm(y ~ x)
summary(mod)
```
:::

## Lists - Why?

```{r, echo=TRUE}
# what is the mod object?
str(mod)
```

## Data Structures

-   vectors

-   factors

-   lists

-   ***`matrices`***

-   arrays

-   dataframes

## Matrices

Matrices are a **two-dimensional** data structure (characterized by 2 dimensions, **`dim()`**), where the number of rows is dimension 1 and the number of columns is dimension 2.

```{r, echo=TRUE}
my_mat = matrix(data = 1:10, nrow = 2, ncol = 5, byrow = FALSE)
my_mat
```

```{r, echo=TRUE}
matrix(data = 1:10, nrow = 2, ncol = 5, byrow = TRUE)
```

## Matrices - Attributes

```{r, echo=TRUE}
attributes(my_mat)
```

```{r, echo=TRUE}
my_mat = matrix(data = 1:10, nrow = 2,
                ncol = 5, dimnames = list(c("row1", "row2"),
                                          c("col1", "col1",
                                            "col3","col4","col5")))
attributes(my_mat)
```

## Matrices - Properties

-   They can contain **only one type** of data

-   Since they are **two-dimensional**, we need two position indices (rows and columns) to identify an element

-   They can be seen as a **set** of single **vectors**

## Matrices - Properties

The number of rows and columns does not have to be the same (square matrix), but the size of the matrix must be compatible with the data vector:

```{r, echo=TRUE, warning=TRUE}
matrix(data = 1:10, ncol = 3, nrow = 3)
```

## Matrices

What does R do by default?

```{r, echo=TRUE, warning=FALSE}

matrix(data = 1:10, ncol = 3, nrow = 3)

matrix(data = 1:2, ncol = 3, nrow = 3)

```

**`warnings`**: the function informs us about something potentially problematic, but (more or less!!) everything is fine

## Matrices - Indexing

To identify one or more elements of a matrix we need row and/or column indices separated by a comma, again inside square brackets: **matrix\[row, column\]**

```{r, echo=TRUE}

my_mat
my_mat[1,1]

```

## Matrices - Indexing

We can also select an entire row or column

```{r, echo=TRUE}

my_mat[1,]

my_mat[,1]

```

## Matrices - Indexing

As with vectors, here too we can apply logical indexing:

```{r, echo=TRUE}
my_mat>2

my_mat

# All the elements greater than two
my_mat[my_mat>2]

```

## Vectors and Matrices

Vectors are created with the **`c()`** function and can be concatenated using the same function:

```{r, echo=TRUE}
my_vect1 = c(1:4)
my_vect2 = c(5:10)

my_vect12 = c(my_vect1,my_vect2)
my_vect12
```

## Matrices

Matrices can be joined together with the commands
```{r, echo=TRUE}

my_mat1 = matrix(data = 1:4,nrow = 2, ncol = 2)
my_mat2 = matrix(data = letters[1:4],nrow = 2, ncol = 2)

```

**`cbind()`**

```{r, echo=TRUE}
cbind(my_mat1, my_mat2)

```

**`rbind()`**

```{r, echo=TRUE}

rbind(my_mat1, my_mat2)

```

Do you notice anything "strange"?

## Operations with matrices

As with vectors, mathematical operations can also be applied to matrices:

```{r, echo=TRUE}

my_mat = matrix(data = 1:4,nrow = 2, ncol = 2)

# element-wise
my_mat*my_mat

# Matrix product
my_mat%*%my_mat
# (1*1 + 3*2) , (1*3 + 3*4)
# (2*1 + 4*2) , (2*3 + 4*4)
```

## Data Structures

-   vectors

-   factors

-   lists

-   matrices

-   ***`arrays`***

-   dataframes

## Arrays

Arrays are n-dimensional objects (e.g., if a matrix is a square, an array is a cube). The same properties of matrices apply, scaled to n dimensions; they are created with the `array` command

```{r, echo=TRUE}

my_array = array(data = 1:6, dim = c(2,3,2)) # three-dimensional example
my_array

```

## Arrays - Indexing

Indexing works the same way as with matrices, adding one dimension

```{r, echo=TRUE}

my_array

my_array[1,1,1] # first row, first column, first "slice"

my_array[1,2,1] # first row, second column, first "slice"

```

## Let's practice! {style="text-align: center;"}

<br/> Open this link and keep it open:

<https://etherpad.wikimedia.org/p/arca-corsoR>
