---
title: "Dataframe"
format:
   revealjs:
     margin: .05
     fontsize: 1.9em
     linestretch: 1.4
     mainfont: Helvetica Neue
editor: visual
---

## Data Structures

-   vectors

-   factors

-   lists

-   matrices

-   arrays

-   ***`dataframes`***

## Dataframe

The dataframe is the most "complex", useful and powerful structure in R. Intuitively it is an excel sheet, while from R's point of view it is a type of **list** with some properties/restrictions

-   each element of the list is a **vector** with an **associated name** (aka a column)
-   each list/column must have the same number of elements
-   as a consequence, each row has the same number of elements (**rectangular structure**)

## Creation

Most of the time you will import dataframes rather than create them, but it is important to first understand how they work and how to create/manipulate them. They are created with the `data.frame` command

```{r, echo=TRUE}
# Create a dataframe with 3 columns
my_df = data.frame( col1 = 1:4, col2 = letters[1:4],
                    col3 = rnorm(n = 4, mean = 0, sd = 1))
my_df
```

## Attributes

The **`dataframe`** has both the attributes of a list, i.e. the ***names***, and the attributes of a matrix, i.e. the ***dimensions*** (rows and columns)

```{r, echo=TRUE}
typeof(my_df)
attributes(my_df)
dim(my_df)
```

------------------------------------------------------------------------

As with the other structures we have seen, we can use the functions **`names()`** , **`dim()`**, **`nrow()`**, **`ncol()`**... to get information about the properties of the dataframe. The most useful function is **`str()`**, since it gives us a quick overview of the structure of the dataframe: dimensions, variable types,...

```{r,echo=TRUE}
str(iris) # iris is a dataset available by default in R (?iris)
```

------------------------------------------------------------------------

## Indexing

We can use either square brackets **`[]`**:

```{r, echo = TRUE}
my_df[1]  # extract a 4x1 data.frame
my_df[[1]] # extract the first column of the data.frame
my_df[1,1] # extract the first element of the first column of the data.frame
```

::: fragment
or the dollar sign **`$`**

```{r, echo=TRUE}
my_df$col1 # extract the first column of the data.frame
my_df$col1[1] # extract the first element of the first column of the data.frame

```
:::

## Indexing - Relational operators

One of the most common operations you will deal with is extracting/evaluating a subset of the values in your dataset:

```{r, echo=TRUE}
my_df

# keep only the rows where the values of column1 are greater than 2
my_df[my_df$col1 > 2, ]
my_df[my_df[1] > 2, ]
```

## Examples

```{r}
my_df
```

::: fragment
```{r, echo=TRUE,eval=FALSE}
my_df[my_df$col1 > 2 & my_df$col1 < 4, ]
```
:::

::: fragment
```{r, eval=TRUE}
my_df[my_df$col1 > 2 & my_df$col1 < 4, ]
```
:::

::: fragment
```{r, echo=TRUE,eval=FALSE}
my_df[my_df$col1== 2,  2]
```
:::

::: fragment
```{r, eval=TRUE}
my_df[my_df$col1== 2,  2]
```
:::

::: fragment
```{r, echo=TRUE,eval=FALSE}
my_df[my_df$col1== 2,  "col2"]
```
:::

::: fragment
```{r, eval=TRUE}
my_df[my_df$col1== 2,  "col2"]
```
:::

## Indexing

There are also alternative, more compact ways of indexing. For example, using the **`subset()`** function:

```{r,echo=TRUE}

subset(iris, subset = Species == "setosa" & Petal.Length > 1.7)

```

equivalent to:

```{r,echo=TRUE}

iris[iris$Species == "setosa" & iris$Petal.Length > 1.7,]

```

------------------------------------------------------------------------

We can also select columns rather than rows with the ***`select`*** argument:

```{r, echo=TRUE, eval=FALSE}
subset(iris, select = c(Sepal.Length, Species))
```

```{r, echo=TRUE}
# show the first rows with the head command
head(subset(iris, select = c(Sepal.Length, Species)), n = 3)
```

------------------------------------------------------------------------

We can also combine the two:

```{r,echo=TRUE, eval=FALSE}
subset(iris, subset = Species == "setosa" & Sepal.Length > 4, select = c(Sepal.Length, Species))
```

```{r,echo=TRUE}
head(subset(iris, subset = Species == "setosa" & Sepal.Length > 4, select = c(Sepal.Length, Species)), n = 3)
```

... equivalent to:

```{r,echo=TRUE}

head(iris[iris$Species == "setosa" & iris$Sepal.Length > 4, c("Sepal.Length","Species")],n = 3)
```

## Indexing

Most of the time you will access variables with the **`$`** operator. This command can also be used to create a new variable...

```{r,echo=TRUE}
# create a variable that is the sum of Length and Width
iris$sum = iris$Sepal.Length + iris$Sepal.Width

str(iris)
```

------------------------------------------------------------------------

The same concepts we have seen for vectors apply, so you can both create and modify variables.

```{r, echo=TRUE}
my_df = data.frame(num = 1:4, let = letters[1:4])
my_df

# Modify the num variable by adding 1
my_df$num = my_df$num+1

# Create a third variable made of the num and let variables
my_df$both = paste(my_df$num,my_df$let, sep = "_") # ?paste

str(my_df)
```

------------------------------------------------------------------------

## Combining Dataframes

Since they are similar to matrices, dataframes can be combined with the **`rbind()`** function:

```{r,echo=TRUE, error=TRUE}
# first dataframe
str(my_df)
# create a second dataframe
my_df2 = data.frame(num = 4:7, lett = letters[1:4],
                    both = paste(4:7,letters[1:4], sep = "_"))

str(my_df2)
```

------------------------------------------------------------------------

**Join the two dataframes**

-   The dataframes must have the same number of columns

-   The column names must be identical

```{r,echo=TRUE, error=TRUE}
my_df3 = rbind(my_df,my_df2)

str(my_df)
str(my_df2)
```

------------------------------------------------------------------------

**Fix the names**

```{r,echo=TRUE, error=TRUE}
names(my_df2)
names(my_df)

# I want the names of my_df2 to match the names of my_df
names(my_df2) = names(my_df)

my_df3 = rbind(my_df,my_df2)
str(my_df3)
```

------------------------------------------------------------------------

You may also need to collect different types of data from the same participant, and then combine the collected information...

Dataframe containing the reaction times:

```{r, echo=TRUE}
df_rt = data.frame(subj = factor(rep(c("alice","bob"),each = 400)),
                   cond = factor(rep(c("easy","hard"),
                                     each = 200, times = 2)),
                      rt = c(rlnorm(n = 400, meanlog = -1, sdlog = .25),
                             rlnorm(n = 400, meanlog = -.7, sdlog =.3)))

# equivalent to first creating the variables as vectors and then putting them in a df
subj = factor(rep(c("alice","bob"),each = 400))
cond = factor(rep(c("easy","hard"),each = 200, times = 2))
rt = c(rlnorm(n = 400, meanlog = -1, sdlog = .25),
                             rlnorm(n = 400, meanlog = -.7, sdlog =.3))

df_rt = data.frame(subj, cond, rt)

```

Dataframe containing the age:

```{r, echo=TRUE}
df_age = data.frame(subj = factor(c("alice","bob")), age = c(20,3))

```

------------------------------------------------------------------------

```{r,echo=TRUE}
str(df_rt) # structure of the reaction times dataframe
head(df_rt)

str(df_age) # structure of the age dataframe
head(df_age)
```

------------------------------------------------------------------------

In this case, we can use the **`merge()`** function and/or the **`_join()`** functions:

```{r, echo=TRUE}
df_all_1 = merge(x = df_rt, y = df_age, by="subj")
str(df_all_1)
```

```{r, echo=TRUE, message=FALSE}
library(dplyr) # load the dplyr package

df_all_2 = left_join(x = df_rt, y = df_age, by = c("subj")) # right_join and full_join also exist
str(df_all_2)
```

## Contingency tables

We can use the **`table()`** function both to compute frequencies and to compute contingency tables, for example to see how many trials we have for each subject.

```{r, echo=TRUE}

# keep only rts between .250 and 1.5 seconds
# i.e. the rows that meet this condition
df_rt_clean = df_rt[df_rt$rt > .25 & df_rt$rt < 1.5, ]

# how many trials for each subject
table(df_rt$subj)
table(df_rt_clean$subj)

# how many trials for each subject x condition
table(df_rt_clean$subj, df_rt_clean$cond)

```

## Exporting and importing data

In R we can import data in many different formats; most commonly you will import ***.csv*** or ***.xlsx*** data.

Here, for example, I export the dataframes in two different formats...

```{r, echo=TRUE}
library(readr) # load the readr package
library(writexl) # load the writexl package

write.csv(df_rt, file = "DATA/df_rt.csv",row.names = FALSE)
write_xlsx(df_age, path = "DATA/df_age.xlsx")
```

## Importing

```{r, echo=TRUE}
library(readxl) # load the readxl package

df_rt_impo = read_csv("DATA/df_rt.csv") # uses the readr package

df_age_impo = read_xlsx("DATA/df_age.xlsx")
```

## Checking the structure

What differences do you notice between the two dataframes?

```{r, echo=TRUE, message=FALSE}
str(df_rt) # Data frame we created
str(df_rt_impo) # Data frame we loaded
```

------------------------------------------------------------------------

The dataframe is imported as a tibble, and in this case the ***`factor`*** variables have been imported as ***`character`***. This can be changed either when importing ( **`?read_csv`** ):

```{r,echo=TRUE,message=FALSE}
df_rt_impo1 = read_csv("DATA/df_rt.csv",
                       col_types =
                         list(col_factor(),col_factor(),
                              col_double()))
str(df_rt_impo1)
```

------------------------------------------------------------------------

or afterwards with the **`as.factor()`** command:

```{r,echo=TRUE,message=FALSE}
df_rt_impo$subj = as.factor(df_rt_impo$subj)
df_rt_impo$cond = as.factor(df_rt_impo$cond)
str(df_rt_impo)
```

------------------------------------------------------------------------

equivalent to:

```{r,echo=TRUE,message=FALSE}
df_rt_impo2 = read_csv("DATA/df_rt.csv") %>% #library(tidyverse)
  mutate(subj = factor(subj),
         cond = factor(cond))
str(df_rt_impo2)
```

(just to show you that in ***R*** a problem can be solved in different ways)

------------------------------------------------------------------------

The same concepts apply to reading xlsx files:

```{r, echo=TRUE}
str(df_age_impo)
df_age_impo$subj=as.factor(df_age_impo$subj)
str(df_age_impo)
df_age_impo$subj=factor(df_age_impo$subj)
str(df_age_impo)
```

------------------------------------------------------------------------

The examples I have shown concern data in formats **external** to R, but we can also save and import data in **`R`** formats.

Their main drawback is that they can be read (mainly) only within the R environment.

If we need to work with different software or languages, it is probably better to use a more generic format.

If we have time left, we will also look at this type of format; the material is available at this [link](https://arca-dpss.github.io/course-R/extra/importing_data.html#1_Introduzione) (in Italian).

## Now let's practice! {style="text-align: center;"}

<br/> Open this link and keep it open:

<https://etherpad.wikimedia.org/p/arca-corsoR>
